Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

\documentclass[runningheads,a4paper]{llncs}
\usepackage{times}
\usepackage{listings}
\usepackage{colonequals}
%\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{framed}

@@

@@

%%% I'm trying to avoid bold whenever it can be avoided (except for headings
%%% etc.) --JB
%\newcommand{\terms}{\mathbf{T}}
\newcommand{\terms}{\mathcal{T}}

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\I}{\mathcal{J}} %%% there's no difference, right? --JB
\newcommand{\J}{\mathcal{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcal{W}}

@@

\newcommand{\Sigmalia}{\Sigma_{LIA}}
\newcommand{\extendsig}[1]{\mathcal{E}( #1 )}

@@

\newcommand{\Mo}{\mathbf{I}}

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols (sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate $\teq$ of type $\tau \times \tau \to \Bool$
for each $\tau \in \ssorts{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite$ of type $\Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$.
We write $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that maps
each $\tau \in \ssorts{\Sigma}$ to a non-empty set $\tau^\I$,
the \emph{domain} of $\tau$ in $\I$;
each $f$ of type $\tau_1 \cdots \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$ 
to a total function $f^\I$ of type $\tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where 
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is 
\emph{$T$-satisfiable} (resp., \emph{$T$-unsatisfiable}) 
if it is satisfied by some (resp., no) interpretation in $\Mo$.
Formulas $\varphi$ and $\psi$ are \emph{equisatisfiable} if
$\varphi$ is $T$-satisfiable if and only if $\psi$ is $T$-satisfiable.
For a given signature $\Sigma$
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the sort and function symbols in this theory as \emph{uninterpreted}.

@@

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are well-specified, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp. non-existence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For instance, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is in fact a non-issue, as $s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

>>>>

\documentclass[runningheads,a4paper]{llncs}
\usepackage{mathptmx}
\usepackage{listings}
\usepackage{colonequals}
%\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{framed}

@@

\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}

@@

%%% I'm trying to avoid bold whenever it can be avoided (except for headings
%%% etc.) --JB
%\newcommand{\terms}{\mathbf{T}}
\newcommand{\terms}{\mathcalx{T}}

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{J}} %%% there's no difference, right? --JB
%% (I find \mathcalx{I} ugly, with its two horizontal bars, hence the J)
\newcommand{\J}{\mathcalx{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcalx{W}}

@@

\newcommand{\Sigmalia}{\Sigma_{LIA}}
\newcommand{\extendsig}[1]{\mathcalx{E}( #1 )}

@@

\newcommand{\Mo}{{\mathcal{\!J\!}}}

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate $\teq$ of type $\tau \times \tau \to \Bool$
for each $\tau \in \ssorts{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite$ of type $\Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$ and
$\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each $\tau \in \ssorts{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in $\I$,
and each $\con{f}$ of type $\tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$ 
to a total function $f^\I$ of type $\tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where 
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is 
\emph{$T$-satisfiable} (resp., \emph{$T$-unsatisfiable}) 
if it is satisfied by some (resp., no) interpretation in $\Mo$.
Formulas $\varphi$ and $\psi$ are \emph{equisatisfiable} if
$\varphi$ is $T$-satisfiable if and only if $\psi$ is $T$-satisfiable.
For a given signature $\Sigma$
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the sort and function symbols in this theory as \emph{uninterpreted}.

@@

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are well-specified, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp.\ nonexistence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For instance, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is not an issue, because
$s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.
